---
title: "Complete Workflow for Whole Body Plethysmography Analysis"
author: "Camron Pearce, PhD"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow for Whole Body Plethysmography Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The **plethR** package provides a complete workflow for analyzing respiratory function data from DSI whole body plethysmography (WBP) systems. This vignette demonstrates a typical analysis from raw data import through publication-quality figures.

## Installation
```{r, eval = FALSE}
# Install from GitHub
devtools::install_github("camronp/plethR")
```
```{r}
library(plethR)
```

# Typical Study Design

Consider a mouse infection study with the following design:

- **4 experimental groups**: Control, Low Dose, Medium Dose, High Dose
- **4 mice per group** (16 total)
- **Time series data**: Daily measurements over 7 days
- **Parameters measured**: Breathing frequency (f), tidal volume (TVb), minute ventilation (MVb), enhanced pause (Penh), and others

# Step 1: Data Import

Import your Excel file containing WBP data. Each sheet represents one mouse's time series data.
```{r, eval = FALSE}
# Import data with time cleaning and apnea removal
df_list <- sheets_into_list(
  excel_file = "my_experiment.xlsx",
  clean_time = TRUE,        # Convert time column to Date format
  remove_apnea = TRUE,      # Remove apnea measurement sheets
  date_average = TRUE       # Average measurements within each day
)
```

The `df_list` object is now a list of 16 data frames (one per mouse), each containing daily averaged respiratory parameters.

**Key parameters:**
- `clean_time`: Converts the Time column to Date format for easier analysis
- `remove_apnea`: Excludes apnea-specific measurements (optional)
- `date_average`: Averages multiple measurements per day into daily values
- `keep_metadata`: Set to `TRUE` if you want to retain subject/phase information

# Step 2: Define Experimental Groups

Create labels for your experimental groups:
```{r, eval = FALSE}
# Define group names
groups <- set_group_names("Control", "Low Dose", "Medium Dose", "High Dose")
```

Alternatively, pass names individually without `c()`:
```{r, eval = FALSE}
groups <- set_group_names("Control", "Low Dose", "Medium Dose", "High Dose")
```

# Step 3: Assign Subjects to Groups

Interactively map which sheets (mice) belong to which groups:
```{r, eval = FALSE}
# Interactive assignment
mapping <- assign_groups_interactive(groups, df_list = df_list)
```

This will prompt you to enter sheet numbers for each group:
```
Enter sheet numbers for group 'Control' as a comma-separated list (e.g., 1,3,5): 
1-4

Enter sheet numbers for group 'Low Dose' as a comma-separated list (e.g., 1,3,5): 
5-8

Enter sheet numbers for group 'Medium Dose' as a comma-separated list (e.g., 1,3,5): 
9-12

Enter sheet numbers for group 'High Dose' as a comma-separated list (e.g., 1,3,5): 
13-16
```

**Tips:**
- Use range notation: `1-4` instead of `1,2,3,4`
- Use comma-separated values: `1,3,5,7`
- Combine both: `1-4,7,9-12`

The function returns a named list mapping groups to sheet indices:
```{r, eval = FALSE}
# Example output
# $Control
# [1] 1 2 3 4
# 
# $`Low Dose`
# [1] 5 6 7 8
```

# Step 4: Calculate Group Averages

Compute mean values for each group at each time point:
```{r, eval = FALSE}
# Calculate group means across time
group_avgs <- calculate_group_averages(
  df_list = df_list,
  group_mapping = mapping,
  time_var = "Time",
  summary_fun = mean,
  keep_n = TRUE              # Include sample size per time point
)
```

This returns a list with one data frame per group, each containing the mean respiratory parameters at each time point.

**Options:**
- `summary_fun = median`: Use median instead of mean
- `summary_fun = sd`: Calculate standard deviations
- `keep_n = TRUE`: Adds a column showing number of observations

# Step 5: Visualize Time Series

Create time series plots to visualize how respiratory parameters change over time:
```{r, eval = FALSE}
# Plot specific parameters with smoothing
plots <- plot_wbp_timeseries(
  data = group_avgs,
  parameters = c("f", "TVb", "MVb", "Penh"),
  plot_type = "by_parameter",    # One plot per parameter
  smooth_method = "rolling",     # Apply rolling average
  smooth_window = 3,             # 3-day window
  show_values = FALSE,
  color_palette = "Set1",
  save_plots = TRUE,
  output_dir = "figures/timeseries",
  dpi = 300
)

# View the frequency plot
plots$f
```

**Plot types:**
- `"by_parameter"`: One plot per parameter, all groups on same plot
- `"by_group"`: One plot per group, all parameters as facets
- `"combined"`: Single multi-panel figure

**Smoothing options:**
- `"none"`: No smoothing
- `"rolling"`: Moving average (specify `smooth_window`)
- `"loess"`: LOESS smoothing (specify `smooth_span`)

# Step 6: Calculate Area Under the Curve (AUC)

AUC summarizes the overall respiratory response across the time series:
```{r, eval = FALSE}
# Calculate AUC normalized to Control group
auc_results <- calculate_auc(
  data = group_avgs,
  time_var = "Time",
  normalize_to = "Control",      # Express as fold-change vs Control
  baseline_correct = FALSE,      # Set TRUE to measure change from day 0
  method = "trapezoid"           # Trapezoidal integration
)

# View results
head(auc_results)
```

**Output:**
```
  group          parameter     auc  auc_normalized  fold_change
  Control        f         25000.0      1.000          1.000
  Low Dose       f         27500.0      1.100          1.100
  Medium Dose    f         30000.0      1.200          1.200
  High Dose      f         32500.0      1.300          1.300
```

**Options:**
- `normalize_to`: Reference group for fold-change calculation
- `baseline_correct = TRUE`: Subtracts first time point before AUC calculation (useful for measuring change from baseline)
- `method = "sum"`: Simple sum (faster, appropriate for evenly-spaced data)

# Step 7: Create AUC Bar Plots

Generate publication-quality bar plots with statistics:
```{r, eval = FALSE}
# Create bar plots with custom group order and statistics
auc_plots <- plot_auc_bars(
  auc_results = auc_results,
  parameters = c("f", "TVb", "MVb", "Penh"),
  group_order = c("Control", "Low Dose", "Medium Dose", "High Dose"),
  value_type = "normalized",     # Plot fold-change values
  y_axis_start = "smart",        # Smart scaling for better visualization
  show_values = TRUE,            # Display values on bars
  show_stats = TRUE,             # Add significance stars
  reference_group = "Control",   # Compare all to Control
  color_palette = "Set1",
  save_plots = TRUE,
  output_dir = "figures/auc",
  file_prefix = "auc",
  width = 6,
  height = 4.5,
  dpi = 600                      # High resolution for publication
)

# View the frequency plot
auc_plots$f_normalized
```

**Key features:**
- **Smart y-axis scaling**: Starts at 80% of minimum value to emphasize differences while maintaining context
- **Statistical annotations**: Adds significance stars (*, **, ***, ****)
- **Custom group ordering**: Control the order of bars on x-axis
- **Value labels**: Shows exact numbers on bars

**For traditional zero-based axes:**
```{r, eval = FALSE}
auc_plots <- plot_auc_bars(
  auc_results = auc_results,
  y_axis_start = "zero",         # Traditional zero-based
  show_stats = TRUE,
  reference_group = "Control"
)
```

**Color schemes:**
- `"Set1"`, `"Set2"`, `"Dark2"`: ColorBrewer palettes
- `"viridis"`, `"plasma"`, `"magma"`: Viridis palettes
- `"grayscale"`: Black and white for print journals

# Step 8: Create Heatmaps

Visualize patterns across all parameters and groups:
```{r, eval = FALSE}
# Create hierarchical clustering heatmap
heatmap <- plot_auc_heatmap(
  auc_results = auc_results,
  value_type = "normalized",
  exclude_parameters = c("Comp", "EEP"),  # Remove less informative parameters
  cluster_rows = TRUE,                    # Cluster parameters
  cluster_cols = TRUE,                    # Cluster groups
  color_scheme = "RdBu",                  # Red-Blue diverging
  show_values = TRUE,                     # Display fold-change in cells
  save_plot = TRUE,
  output_dir = "figures/heatmaps",
  dpi = 300
)
```

**Clustering methods:**
- `"complete"`: Complete linkage (default)
- `"average"`: Average linkage
- `"ward.D2"`: Ward's method

**Color schemes:**
- `"RdYlBu"`: Red-Yellow-Blue (default for normalized data)
- `"RdBu"`: Red-Blue diverging
- `"PRGn"`, `"BrBG"`: Colorblind-friendly diverging
- `"viridis"`, `"plasma"`: Sequential

**Include only specific parameters:**
```{r, eval = FALSE}
heatmap <- plot_auc_heatmap(
  auc_results = auc_results,
  include_parameters = c("f", "TVb", "MVb", "Penh", "PAU")
)
```

# Step 9: Principal Component Analysis (PCA)

PCA reveals overall patterns and identifies which mice are similar:
```{r, eval = FALSE}
# PCA with loading vectors showing which parameters drive separation
pca_result <- plot_pca(
  data = df_list,
  group_mapping = mapping,
  show_loadings = TRUE,          # Show which parameters contribute most
  n_loadings = 5,                # Display top 5 contributors
  show_ellipses = TRUE,          # 95% confidence ellipses
  color_palette = "Set1",
  save_plot = TRUE,
  output_dir = "figures/pca",
  dpi = 300
)

# View the plot
pca_result$plot

# Check variance explained
print(pca_result$variance)

# View PC scores
head(pca_result$scores)
```

**Output:**
- Each point represents one mouse
- Distance between points indicates similarity in respiratory patterns
- Loading vectors show which parameters (f, TVb, etc.) drive the separation

**Check variance explained:**
```{r, eval = FALSE}
# PC1 explains 48.6% of variance
# PC2 explains 29.4% of variance
# Cumulative: 78.0%
```

**With k-means clustering:**
```{r, eval = FALSE}
# Apply unsupervised clustering
pca_cluster <- plot_pca(
  data = df_list,
  group_mapping = mapping,
  use_clustering = TRUE,
  num_clusters = 4,
  color_by = "cluster",          # Color by clusters instead of groups
  save_plot = TRUE
)

# Compare clusters to experimental groups
table(Experimental = pca_cluster$scores$group, 
      Cluster = pca_cluster$scores$cluster)
```

**Plot different principal components:**
```{r, eval = FALSE}
# PC3 vs PC4
pca_34 <- plot_pca(
  data = df_list,
  group_mapping = mapping,
  components = c("PC3", "PC4")
)
```

# Step 10: Export Processed Data

Save your results for collaborators or further analysis:
```{r, eval = FALSE}
# Export group averages
export_to_excel(
  data_frames = group_avgs,
  file_name = "group_averages",
  output_dir = "results"
)

# Export AUC results
export_to_excel(
  data_frames = auc_results,
  file_name = "auc_analysis",
  output_dir = "results"
)

# Export with timestamp to avoid overwriting
export_to_excel(
  data_frames = group_avgs,
  file_name = "group_averages",
  timestamp = TRUE              # Adds date-time to filename
)
```

# Complete Example Script

Here's a complete analysis script you can adapt:
```{r, eval = FALSE}
library(plethR)

# 1. Import data
df_list <- sheets_into_list("experiment.xlsx", 
                             clean_time = TRUE, 
                             date_average = TRUE)

# 2. Define and assign groups
groups <- set_group_names("Control", "Low", "Medium", "High")
mapping <- assign_groups_interactive(groups, df_list)

# 3. Calculate group averages
averages <- calculate_group_averages(df_list, mapping)

# 4. Visualize time series
ts_plots <- plot_wbp_timeseries(
  averages,
  parameters = c("f", "TVb", "Penh"),
  smooth_method = "rolling",
  smooth_window = 3,
  save_plots = TRUE
)

# 5. Calculate AUC
auc <- calculate_auc(averages, normalize_to = "Control")

# 6. Create bar plots
bar_plots <- plot_auc_bars(
  auc,
  group_order = groups,
  show_stats = TRUE,
  reference_group = "Control",
  save_plots = TRUE,
  dpi = 600
)

# 7. Create heatmap
heatmap <- plot_auc_heatmap(
  auc,
  include_parameters = c("f", "TVb", "MVb", "Penh"),
  color_scheme = "RdBu",
  save_plot = TRUE
)

# 8. PCA analysis
pca <- plot_pca(
  df_list,
  group_mapping = mapping,
  show_loadings = TRUE,
  save_plot = TRUE
)

# 9. Export results
export_to_excel(averages, "group_averages", output_dir = "results")
export_to_excel(auc, "auc_results", output_dir = "results")

# View variance explained by PCA
print(pca$variance)
```

# Tips for Publication-Quality Figures

1. **Use high DPI for journals**: Set `dpi = 600` for submission-ready figures

2. **Maintain consistent dimensions**: Use the same `width` and `height` across related figures

3. **Smart y-axis scaling**: Use `y_axis_start = "smart"` to emphasize differences, but consider `y_axis_start = "zero"` if journal requires it

4. **Logical group ordering**: Order groups from control to highest treatment with `group_order`

5. **Colorblind-friendly palettes**: Use `"Set2"`, `"Dark2"`, or `"PRGn"` for accessibility

6. **Statistical annotations**: Always specify `reference_group` when using `show_stats = TRUE`

7. **Consistent color schemes**: Use the same palette across all figures in a publication

# Common Workflows

## Infection Study (Pre/Post infection)
```{r, eval = FALSE}
# Calculate baseline-corrected AUC
auc <- calculate_auc(
  averages,
  baseline_correct = TRUE,      # Measures change from day 0
  normalize_to = "Uninfected"
)
```

## Dose-Response Study
```{r, eval = FALSE}
# Order groups by increasing dose
bar_plots <- plot_auc_bars(
  auc,
  group_order = c("Vehicle", "0.1mg", "1mg", "10mg", "100mg"),
  show_stats = TRUE,
  reference_group = "Vehicle"
)
```

## Genotype Comparison
```{r, eval = FALSE}
# Group organization
groups <- set_group_names("WT Baseline", "KO Baseline", 
                         "WT Treatment", "KO Treatment")

# Logical ordering for visualization
plot_auc_bars(
  auc,
  group_order = c("WT Baseline", "WT Treatment", 
                  "KO Baseline", "KO Treatment")
)
```

# Getting Help

- Package documentation: `?plethR`
- Function help: `?function_name` (e.g., `?calculate_auc`)
- GitHub issues: https://github.com/camronp/plethR/issues
- List all functions: `ls("package:plethR")`

# Session Info
```{r}
sessionInfo()
```